---
title: Homepage
author: Eric Bryant, PhD
include-before: >
  First time here? Check out the [FAQ](/faq/)! Otherwise, jump to:
output:
  blogdown::html_page:
    toc: yes
---


**March 13^th^ 2021 note:** CA deprecated the links to the original data sources
for this tracker. It appears the numbers have been changed (likely some cleaning
of historical data has occurred because a few anomalies are now gone).
I will ad a News entry when I get a chance to review the changes in more detail.
In the meantime, I have migrated to the new source of new case information and
will migrate to the new source of hospitalization data soon. 


```{r include = FALSE}
# Load R packages, functions, and settings
source(here::here("requirements.R"))

# Where to save new cases in CA
csv_cases <- here("data/new-cases-ca-v2.csv")
csv_hsp   <- here("data/hospitalization-ca.csv")

# ---- !!! Population data now provided from CA directly as of 2021-03-13 ----
# Population from: https://worldpopulationreview.com/us-counties/ca/los-angeles-county-population
#la_population <- 10079000

# Population from: https://worldpopulationreview.com/us-counties/ca/santa-clara-county-population
#sc_population <- 1945940
# ---- !!! -------------------------------------------------------------------

# Range of dates
dates <- lubridate::today("UTC") + c(-255, 110)

# Download data from https://data.ca.gov
# Only downloads once a day unless force = TRUE
download_new_cases_ca_csv_v2(csv_cases)
#download_hospitalization_ca_csv(csv_hsp)
```


Los Angeles
--------------------------------------------------------------------------------

For the most recent information, I recommend reading the
[LA county public health press releases](http://publichealth.lacounty.gov/media/coronavirus/#news).

```{r covid-los-angeles, fig.height = 7}
county <- "Los Angeles"

la_population <-
  read_csv(csv_cases, col_types = cols()) %>%
  filter(county == {{ county }}) %>%
  pull(population) %>%
  median() # compromise since second scale needs to be a linear transformation

figure_new_cases(csv_cases, county, population = la_population, date_limits = dates) /
  figure_new_cases_change(csv_cases, county, date_limits = dates) #/
#  figure_hospitalization(csv_hsp, county, date_limits = dates)
```


Santa Clara
--------------------------------------------------------------------------------

For the most recent information, I recommend reading the
[Santa Clara county public health press releases](https://www.sccgov.org/sites/covid19/Pages/archives.aspx).

```{r covid-santa-clara, fig.height = 7}
county <- "Santa Clara"

sc_population <-
  read_csv(csv_cases, col_types = cols()) %>%
  filter(county == {{ county }}) %>%
  pull(population) %>%
  median() # compromise since second scale needs to be a linear transformation

figure_new_cases(csv_cases, county, population = sc_population, date_limits = dates) /
  figure_new_cases_change(csv_cases, county, date_limits = dates) #/ 
  #figure_hospitalization(csv_hsp, county, date_limits = dates)
```
