---
title: Homepage
author: Eric Bryant, PhD
output:
  blogdown::html_page:
    toc: no
---



```{r include = FALSE}
# Load R packages, functions, and settings
source(here::here("requirements.R"))

# Where to save new cases in CA
csv_cases <- here("data/new-cases-ca-v2.csv")
csv_hsp   <- here("data/hospitalization-ca-v2.csv")

# ---- !!! Population data now provided from CA directly as of 2021-03-13 ----
# Population from: https://worldpopulationreview.com/us-counties/ca/los-angeles-county-population
#la_population <- 10079000

# Population from: https://worldpopulationreview.com/us-counties/ca/santa-clara-county-population
#sc_population <- 1945940
# ---- !!! -------------------------------------------------------------------

# Range of dates
#dates <- lubridate::today("UTC") + c(-255, 110)
#dates <- lubridate::today("UTC") + c(-500, 150)
dates <- c(as.Date("2020-11-01"), lubridate::today("UTC") + 150)

# Download data from https://data.ca.gov
# Only downloads once a day unless force = TRUE
download_new_cases_ca_csv_v2(csv_cases)
download_hospitalization_ca_csv_v2(csv_hsp)
```


For the most recent information, read the
[LA county public health press releases](http://publichealth.lacounty.gov/media/coronavirus/#news).

```{r covid-los-angeles, fig.height = 7}
county <- "Los Angeles"

la_population <-
  read_csv(csv_cases, col_types = cols()) %>%
  filter(county == {{ county }}) %>%
  pull(population) %>%
  median() # compromise since second scale needs to be a linear transformation

figure_new_cases(csv_cases, county, population = la_population, date_limits = dates) /
  figure_new_cases_change(csv_cases, county, date_limits = dates) /
  figure_hospitalization(csv_hsp, county, date_limits = dates)
```


Below are the reported numbers for the last 30 days. Use this to check for
reporting weirdness and to get a more precise feel of LA numbers over time.

```{r covid-los-angeles-table-01}
df_cases <-
  read_csv(csv_cases, col_types = cols()) %>% 
  filter(.data$county == .env$county) %>%
  # Calculate estimates
  arrange(date) %>%
  mutate(
    count_new_roll  = slide_dbl(count_new, median, .before = 7L, .after = 7L),
    count_new_loess = predict_loess(date, count_new, span = 0.11),
    count_new_loess_diff = count_new_loess - lag(count_new_loess),
    count_new_loess_fc   = count_new_loess / lag(count_new_loess),
    count_new_loess_diff_weekly = count_new_loess - lag(count_new_loess, 7L),
    count_new_loess_fc_weekly = count_new_loess / lag(count_new_loess, 7L),
    death_new_loess = predict_loess(date, death_new, span = 0.1)
  ) %>%
  select(
    Date = date,
    "New cases" = count_new,
    "Smooth estimate" = count_new_loess,
    "2 week mean" = count_new_roll,
    "Daily change" = count_new_loess_diff,
    "Weekly change" = count_new_loess_diff_weekly,
    "Daily fold change" = count_new_loess_fc,
    "Weekly fold change" = count_new_loess_fc_weekly,
    "New deaths" = death_new,
    "Death estimate" = death_new_loess
  )

df_hsp <-
  read_csv(csv_hsp, col_types = cols()) %>%
  # Date hard censored due to change in reporting criteria
  filter(.data$county == .env$county, date >= as.Date("2020-05-01")) %>%
  # Combine confirmed and suspected counts (at this point most are confirmed)
  # In the spring there were more "suspected" probably due to testing backlogs?
  mutate(
    hsp = hsp_covid_confirmed + hsp_covid_suspected,
    icu = icu_covid_confirmed + icu_covid_suspected,
    hsp = hsp - icu # remove ICU cases from inpatient (the figure will stack)
  ) %>%
  select(Date = date, "Hospitalized" = hsp, "ICU" = icu, "Open ICU beds" = icu_available_beds)

df_data <- left_join(df_hsp, df_cases, by = "Date")

df_data %>%
  top_n(30, Date) %>%
  arrange(desc(Date)) %>%
  #mutate(across(where(is_double) & !Date, ~as.integer(round(.)))) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(`Hospitalized`, ICU, `New deaths`, `New cases`, `Death estimate`, `Smooth estimate`, `2 week mean`, `Daily change`, `Weekly change`), decimal = 0) %>%
  gt::fmt_number(columns = ends_with("fold change"), decimal = 2) %>%
  gt::cols_width(
    Date ~ px(100),
    ICU  ~ px(80)
  ) %>%
  gt::gtsave(paste0(tempfile(), ".png"))
```


## Playground

Below is a comparison of cases, hospitalizations, ICU and deaths relative to the
peak in winter 2021. Hospitalizations and ICU admissions have been shifted back
9 days, and Deaths have been shifted back 22 days as this was the approximate
lag for the winter 2021 wave.

```{r covid-los-angeles-figure-02, fig.height = 2.5}
x <-
  df_data %>%
  filter(Date > as.Date("2020-09-01")) %>%
  #filter(complete.cases(.)) %>%
  mutate(
    norm_cases = 
      (`2 week mean` - min(`2 week mean`, na.rm = TRUE)) / 
      max(`2 week mean`[Date < as.Date("2021-04-01")] - min(`2 week mean`, na.rm = TRUE), na.rm = TRUE),
    norm_hsp   =
      (Hospitalized - min(`Hospitalized`, na.rm = TRUE)) /
      max(Hospitalized[Date < as.Date("2021-04-01")] - min(Hospitalized, na.rm = TRUE), na.rm = TRUE),
    norm_icu   =
      (ICU - min(ICU, na.rm = TRUE)) /
      max(ICU[Date < as.Date("2021-04-01")] - min(ICU, na.rm = TRUE), na.rm = TRUE),
    norm_deaths =
      (`Death estimate` - min(`Death estimate`, na.rm = TRUE)) /
      max(`Death estimate`[Date < as.Date("2021-08-01")] - min(`Death estimate`, na.rm = TRUE), na.rm = TRUE),
    lag_adjusted_hsp = lead(norm_hsp, 9),
    lag_adjusted_icu = lead(norm_icu, 9),
    lag_adjusted_deaths = lead(norm_deaths, 22)
  )

x %>%
  select(
    Date,
    Cases = norm_cases,
    `Lag adjusted hospitalizations` = lag_adjusted_hsp,
    `Lag adjusted ICU` = lag_adjusted_icu,
    `Lag adjusted Deaths` = lag_adjusted_deaths
  ) %>%
  pivot_longer(cols = -Date, names_to = "group", values_to = "Normalized counts") %>%
  filter(complete.cases(.)) %>%
  ggplot(aes(x = Date, y = `Normalized counts`, color = group)) +
  scale_color_manual(values = c("red", "gold2", "#00a174", "#ec5500")) +
  scale_x_date(
    date_breaks = "4 month",
    date_labels = "%b",
    date_minor_breaks = "1 month",
    sec.axis =
      dup_axis(
        breaks = scales::date_breaks("1 year"),
        labels = scales::date_format("%Y")
      )
  ) +
  scale_y_continuous(breaks = c(0, 0.25, 0.5, 0.75, 1, 2, 3, 4), labels = scales::percent) +
  geom_vline(xintercept = as.Date(c("2021-01-01", "2022-01-01", "2023-01-01"), tz = ""), linetype = "dotted") +
  geom_line(size = 0.2) +
  labs(
    title = str_c(county, " county"),
    y = "Normalized counts",
    caption =
      str_c(
        "<span style='color:red'>Red line</span>: New cases (2 week rolling average)",
        "<span style='color:#00a174'>Green line</span>: Hospitalizations, 9 day lag adjusted",
        "<span style='color:#ec5500'>Orange line</span>: ICU, 9 day lag adjusted",
        "<span style='color:gold2'>Yellow line</span>: Deaths estimate, 22 day lag adjusted",
        sep = "<br>"
      )
  ) +
  theme(
    legend.position = "none",
    plot.caption = ggtext::element_markdown(hjust = 0, size = 7),
    aspect.ratio = 0.5,
    axis.title.x = element_blank(),
    axis.text.x = element_text(hjust = 0)
  )

```

```{r echo = FALSE}
n_omicron_cases <-
  df_cases %>%
  filter(Date >= as.Date("2021-12-01")) %>%
  pull(`New cases`) %>%
  sum()

frac_omicron_cases <- n_omicron_cases / la_population

chr_n_omicron_cases <- scales::number(n_omicron_cases, big.mark = ",")
chr_pct_omicron_cases <- scales::percent(frac_omicron_cases, accuracy = 0.1)
```

With so many new cases in recent days, I was curious what fraction of LA
county's population has gotten covid during the omicron wave. Without
considering duplicates, and assuming that December 2021 was the start of this
wave, there have been `r chr_n_omicron_cases` confirmed cases, which is roughly
`r chr_pct_omicron_cases` of the population.

Saw some chatter on Twitter about the circular plot of cases published by the
NYT
[here behind a paywall](https://www.nytimes.com/2022/01/06/opinion/omicron-covid-us.html).
My personal take is that the figure evokes more emotion than it provides useful
information, which I think is the point since this is a NYT Opinion piece. 
Anyhow, I thought it might be fun to try to reproduce the figure with LA data.
(now with code!)

```{r covid-los-angeles-figure-03, echo = TRUE}
# All of this just to have grid lines that vary in length for each month!
df_segments <-
  tibble(
    x = seq(0, 360, by = 364 / 12),
    yend = c(1000, seq(500, 950, by = 500 / 12)),
    month = month.abb,
    hjust = c(0.5, -0.1, -0.1, -0.1, -0.1, -0.1, 0.5, 1.1, 1.1, 1.1, 1.1, 1.1),
    vjust = c(-0.1, -0.1, -0.1, 0.5, 1.1, 1.1, 1.1, 1.1, 1.1, 0.5, -0.1, -0.1)
  )

# Prepare the plot data
df_plot <-
  df_cases |>
  # Only need date and cases
  select(
    Date,
    Cases = `Smooth estimate` # Using smooth estimate, cause smooth is nice
  ) |>
  # Remove the one day in the leap year that makes things too hard *sadface*
  filter(as.character(Date) != "2020-02-29") |>
  # Going to put the dates in order so I can enumerate them to create a spiral
  # (i.e. increasing radius as we go around the circle)
  arrange(Date) |>
  mutate(
    # Sometimes the smoother goes below 0, we'll just say those are 0
    cases = ifelse(Cases < 0, 0, Cases),
    # To mimic the times choice of blobbing out, I'll divide by 2
    cases_half = cases / 2,
    # I then rescale everything so that they, well, fit without overlapping
    cases_half = cases_half / 85, # trial and error to get this number
    #day_of_year = (lubridate::yday(Date) - 0.5),  * 0.9875,
    day_of_year = lubridate::yday(Date) - 0.5,
    year = as.factor(lubridate::year(Date)),
    spiral = 1:n()
  )

# Given or plot data
df_plot |>
  # I plot the day of the year (1 to 365), and I group cases by year to 
  # allow for plotting one ribbon multiple times to a given x value
  ggplot(aes(x = day_of_year, group = year)) +
  # I add my ribbon
  geom_ribbon(
    # With a base at the spiral and the ribon spanning equally above and below
    # the spiral
    aes(y = spiral, ymax = spiral + cases_half, ymin = spiral - cases_half),
    # I make it pretty and try to match the color scheme from NYT
    alpha = 1,
    color = "#ce0f10",
    fill = "#f9e0df"
  ) +
  # I manually add grid lines to allow them to vary in length
  geom_segment(
    aes(x = x, xend = x, y = 0, yend = yend),
    linetype = "dashed",
    color = "grey",
    alpha = 0.5,
    data = df_segments,
    inherit.aes = FALSE
  ) +
  # Add manually add the month labels since they vary in position
  geom_text(
    aes(x = x, y = yend, label = month,  hjust = hjust, vjust = vjust),
    inherit.aes = FALSE,
    data = df_segments
  ) +
  # I make some adjustments so the yearly period lines up to the circle
  scale_x_continuous(
    limits = c(0, 365),
    expand = c(0, 0)
  ) +
  # I make the plot a circle
  coord_polar() +
  # Give it a title and remove theme elements I no longer need
  labs(title = "New Covid-19 cases,\nLos Angeles County") +
  theme(
    axis.title = element_blank(),
    panel.grid = element_blank(),
    axis.text = element_blank()
  )
```


