---
title: Hospitalizations
author: Eric Bryant, PhD
date: '2020-11-25'
slug: hospitalizations
categories: []
tags: []
---


With cases rapidly rising in California, I have decided to add daily
hospitalization updates to the homepage. This post outlines the "making of"
the hospitalization figure.

The data are described in detail
[here](https://data.ca.gov/dataset/covid-19-hospital-data/resource/42d33765-20fd-44b8-a978-b083b7542225).
If you really want to understand the data, this is where you should start.

I begin by reading the data and focusing on Los Angeles. I am only going to look
at data starting in May, since there is a note on
[data.ca.gov](https://data.ca.gov/dataset/covid-19-hospital-data/resource/42d33765-20fd-44b8-a978-b083b7542225)
saying that there was a change in reporting at the end of April (namely that
emergency departments are no longer included in these figures). Also note that
I changed the column names during download to be a little be more consistent
(just me being persnickety).

```{r message = FALSE}
source(here::here("requirements.R"))

df <-
  read_csv(here("data/hospitalization-ca.csv")) %>%
  filter(county == "Los Angeles", date >= as.Date("2020-05-01"))
```


First I just want to get a sense of the data, so I make a table with the median
and range of data points found in each column.

```{r echo = TRUE}
df %>%
  select(-county) %>% # drop the county column (it's all just "Los Angeles")
  gtsummary::tbl_summary(
    # Report the range instead of the default IQR
    statistic = list(gtsummary::all_continuous() ~ "{median} [{min} - {max}]")
  )
```

Here we see that we have daily numbers for ICU and hospitalized patients that
are either confirmed or suspected to have COVID-19. I think these numbers are
pretty self-explanatory, but there are a couple of important details:

1. Hospitalized patients are occupying inpatient beds which **includes** ICU and
   surgical units, but does **not** include overflow and emergency department
   beds.
2. All ICU numbers **include** NICU (neonatal) and PICU (pediatric) beds.
3. Reported available beds are not necessarily staffed.

So, what do these numbers look like over time?

```{r hospitalizations, echo = TRUE, fig.height = 3.5}
# Prepare the data for plotting
plot_data <-
  df %>%
  # Combine confirmed and suspected counts (at this point most are confirmed)
  # In the spring there were more "suspected" probably due to testing backlogs?
  mutate(
    hsp = hsp_covid_confirmed + hsp_covid_suspected,
    icu = icu_covid_confirmed + icu_covid_suspected,
    hsp = hsp - icu # remove ICU cases from inpatient (the figure will stack)
  ) %>%
  # Reshape into a long table to make plotting easier
  select(date, hsp, icu) %>%
  pivot_longer(c("hsp", "icu"), names_to = "type", values_to = "count") %>%
  mutate(
    type =
      factor(type, levels = c("hsp", "icu"), labels = c("Inpatient", "ICU"))
  )

# Plot date vs count and color by ICU/inpatient
gg_hsp_icu <-
  ggplot(plot_data, aes(date, count, color = type, fill = type)) +
  geom_area(alpha = 0.5) # Numbers get stacked

gg_hsp_icu
```


This is our basic plot without any customization. It is good enough for casual
inspection -- we can see our peak COVID hospitalizations happened in July,
during the last peak of new cases and is currently rising rapidly.

Before adding this figure to the site, I need to clean it up to make it more
readable and consistent with the format of the other timeline figures. As you
can see, making things pretty takes a lot more effort.

```{r hospitalizations-pretty, echo = TRUE, fig.height = 3.5}
# Match the date range of our other figures
date_limits <- c("2020-03-17", "2021-03-01")

# All this just makes the plot pretty
gg_hsp_icu +
  # Add a line to mark the new year
  geom_vline(xintercept = as.Date("2021-01-01", tz = ""), linetype = "dotted") +
  labs(
    # be clear about the numbers in the figure caption
    caption = str_c(
      "COVID confirmed + suspected",
      "Numbers do not include ER, overflow, or outpatient",
      sep = "<br>"
    ),
    y = "Patients"
  ) + 
  scale_fill_brewer(palette = "Dark2") +       # Pick different colors
  scale_color_brewer(palette = "Dark2") +
  scale_y_continuous(labels = scales::comma) + # Add commas to the numbers
  scale_x_date(
    date_breaks = "2 month",       # Label the date every two months
    date_labels = "%b",            # Label with just the month abbreviation
    date_minor_breaks = "1 month", # add vertical lines for every month
    limits = as.Date(date_limits, tz = ""),     # set the date limits
    sec.axis =
      dup_axis(
        breaks = scales::date_breaks("1 year"), # Add a 1 year marker
        labels = scales::date_format("%Y")      # Label the year
      )
  ) +
  theme(
    # Ugh, legends are annoying
    legend.title = element_blank(),
    legend.key.height = unit(3, "points"),
    legend.key.width  = unit(25, "points"),
    legend.text = element_text(size = 6),
    legend.direction = "horizontal",
    legend.position = "bottom",
    legend.box.spacing = unit(1, "points"),
    plot.caption = ggtext::element_markdown(hjust = 0, size = 7),
    aspect.ratio = 0.5,
    axis.title.x = element_blank(),
    axis.text.x = element_text(hjust = 0)
  )
```


Ok, that looks good to my eye (well actually it looks bad, that rise in patients
is very steep).

Now, what about those available bed numbers?

```{r hospitalizations-beds, echo = TRUE, fig.height = 3.5}
gg_beds <-
  df %>%
  ggplot(aes(date, hsp_all_beds)) +
  geom_line() +
  coord_cartesian(ylim = c(0, NA))

gg_icu_beds <-
  df %>%
  ggplot(aes(date, icu_available_beds)) +
  geom_line() +
  coord_cartesian(ylim = c(0, NA))

gg_beds / gg_icu_beds
```

The reported number of beds doesn't seem to change much, and there is a big drop
in August that I can't find details about. The take-home here for "all beds" is
that there are roughly 20,000 hospital beds in LA county.

As for the ICU available beds, I think this number reports how many beds are
unoccupied? Unfortunately I can't really confirm this and the number of
"available" beds doesn't track well with the ICU patient numbers, so I am not
sure what to make of this other than there are roughly 1,000 ICU beds in LA
county (including NICU and PICU). If I am reading this right though, we are at
our current lowest available ICU bed capacity.

These numbers are weird enough that I don't feel confident adding them to the
tracker -- I have too many questions about the random drops and rises.

Thoughts, questions, concerns, suggestions? Message me on Twitter
[\@ericedwrdbryant](https://twitter.com/ericedwrdbryant).
