---
title: Why are daily new cases so wiggly?
author: Eric Bryant, PhD
date: '2020-11-20'
slug: why-are-daily-new-cases-so-wiggly
categories: []
tags: []
---

What's the deal with all the wiggliness in daily new cases? In the FAQ I say
this is a reporting artifact because it looks like it happens at a regular
interval approximately once a week, but I have never really checked this
closely. 

Do the dips that we see in daily new cases happen on particular days of the
week? Let's have a look! (BTW I'm leaving the code with lots of comments so you
can follow along if that kind of thing is interesting to you).

I begin by loading all the R packages I intend to use (really just the
[tidyverse](https://www.tidyverse.org) for this analysis, plus
[gtsummary](http://www.danieldsjoberg.com/gtsummary/) -- Hi Margie!). Then I
read in the daily new case data for Los Angeles.

```{r message = FALSE}
# Load R packages and project settings (go to GitHub for contents of this file)
source(here::here("requirements.R"))

df <-
  read_csv(here("data/new-cases-ca.csv")) %>% # Read the data
  filter(county == "Los Angeles") %>%         # We'll focus on LA
  mutate(
    # Get the day of the week given the date
    weekday = lubridate::wday(date, label = TRUE, week_start = 1),
    # Label weekends for color in our plot
    Weekend = weekday %in% c("Sat", "Sun"),
    # Calculate the deviation from expected new cases
    count_new_diff = count_new - count_new_roll
  )
```


Now that we have our data, I make a boxplot to get an idea for how the deviation
of expected new cases is distributed on each week day.

```{r}
# Given our data
df %>%
  # Plot weekday vs. expected new cases. Color by weekend (yes/no)
  ggplot(aes(x = weekday, y = count_new_diff, color = Weekend)) +
  labs(y = "Deviation from expected new cases") +
  geom_boxplot(outlier.shape = NA) +       # don't worry about outliers
  scale_color_brewer(palette = "Dark2") +  # use pretty colors
  theme(aspect.ratio = 1) +                # make plot region a square
  coord_cartesian(ylim = c(-1000, 1000)) + # Set limits on Y-axis
  theme(
    legend.position = "none",              # no need for color legend
    axis.title.x = element_blank()         # no need for X axis title
  )
```


Looks like numbers drop on weekends! Why? I don't know, but I assume there is
just limited reporting capacity when people take the weekend off?

OK, how big is this effect? Well, let's compare the average deviation from
expected new cases for each day of the week to Wednesday (since it's in the
middle of the week I figure it should be furthest away from the weekend drop)

```{r}
# Given our data
df %>%
  # Set Wednesday as the "reference" since it's in the middle of the week
  mutate(Weekday = relevel(factor(weekday, ordered = FALSE), ref = "Wed")) %>%
  lm(count_new_diff ~ Weekday, data = .) %>% # fit a linear model
  gtsummary::tbl_regression()                # make a pretty table
```

And there it is, on average we see roughly 300 to 400 fewer cases on Saturdays
and Sundays relative to Wednesdays. That means we under-report cases on weekends
to the tune of roughly 700 new cases. I would guess these cases get reported
later in the week, but I don't really know.

**ONE MORE IMPORTANT THING!** I look at the data every day when it comes in and
I don't remember the drops happening on weekends. It turns out that the numbers
I download from [data.ca.gov](https://data.ca.gov) are not for that day, or even
the day before. Rather, Saturday's numbers become available in this database
around 12pm on Monday, and Sunday's numbers show up on Tuesday -- a 2 day delay.

All that said, you should expect the **daily updates on this site to be two days
old and artificially low on Mondays and Tuesdays.**
