---
title: LA county restrictions
author: Eric Bryant, PhD
date: '2020-11-23'
slug: la-county-restrictions
categories: []
tags: []
---

**Update as of 5pm Mon, Nov 23 2020:**
_[Some reports](https://laist.com/latest/post/20201123/la-county-covid-stay-at-home-curfew-order)
have indicated that Monday's numbers hit 6,124 and that this is triggering a
three week stay-at-home. This is sort of surprising to me since, if these are
Monday's numbers, then Sunday would have to be over 5,500. Or, perhaps these
"Monday" numbers are Sunday's numbers (which would make sense, since Monday is
not over yet). This would make the five day average 4,358. While this is not
over the 4,500 threshold, this could be explained away by the small
discrepancies I see between LACPH and my data source
[data.ca.gov](https://data.ca.gov)). Either way, looks like stay-at-home might
be called much sooner than expected?_

Due to recent increases in daily new cases in Los Angeles,
[LA county has issued a heads-up](https://covid19.lacounty.gov/covid19-news/los-angeles-county-to-implement-tighter-safeguards-and-restrictions-to-curb-covid-19-spread/)
that new restrictions may be put into place in the coming days/weeks.

I'll summarize some relevant new measures here, but do look at the press release
for complete details (and in-case I make a typo here). And for that matter,
it's not a bad idea to check in on the latest
[LA county news releases](https://covid19.lacounty.gov/covid19-news).

As of last Friday:

- Non-essential indoor work spaces limited to 25% capacity.
- Restaurants etc. are limited to 50% outdoor capacity, must close 10pm to 6am.
- Indoor services by appointment only, mask required (so no facials or shaves).
- 15 person, 3 household maximum for outdoor gatherings (no indoor gatherings). 

If the **5 day** average of new cases is **>4,000**, or hospitalizations is
**>1,750**:

- Take out dining only.

If the **5 day** average of new cases is **>4,500** or hospitalizations is
**>2,000**:

- ***Stay at home!*** Essential work and services only. 10pm to 6am curfew.


## Where are we now?

First I would like to note that data from LA county public health, which is
linked at the bottom of the press-release has some discrepancies
(generally small) with county level data reported to the state. I don't know why
this is, but just be aware that official decision makers are probably using
LACPH data rather than the CA state data that most trackers use -- including
this site.

Our current numbers for the last 5 days from CA state are:

```{r message = FALSE}
source(here::here("requirements.R"))

la_5_day <-
  read_csv(here("data/new-cases-ca.csv")) %>%
  filter(county == "Los Angeles") %>%
  mutate(weekday = lubridate::wday(date, label = TRUE)) %>%
  select(date, weekday, count_new, count_new_roll, count_new_loess) %>%
  top_n(5, date) %>%
  print()
```

The average for these last 5 days is
`r format(round(mean(la_5_day$count_new)), big.mark = ",")`. Note that the latest
reported number is from a Saturday where we see the expected artificial drop.

Will a stay-at-home order be issued? Depends on how bad things are on Monday and
Tuesday. For stay-at-home to be issued on Wednesday night, we need to know what
the numbers will have to average out to on Monday and Tuesday. We don't have
Sunday's numbers but we can sort of guess that it will be artificially low like
Saturday (big uncertainty here though).

```{r echo = TRUE, comment = ""}
# I'm going to just guess that Sunday will be artificially low to the same
# degree as Saturday
#             Threshold     Fri    Sat    Sun
mt_thresh <- ((4500 * 5) - (4200 + 2587 + 2600)) / 2
cat(mt_thresh)
```

If we take last Friday's numbers and we assume the daily rate of change:

```{r echo = TRUE, comment = ""}
mt_guess <- mean(c(4200 + (120 + 240 + 360), 4200 + (120 + 240 + 360 + 480)))
cat(mt_guess)
```

That would still leave us about
`r format(round(mt_thresh - mt_guess, digits = -2), big.mark = ",")` cases under
what would be required to call a lockdown.

**Important caveats:**

1. This is dumb napkin math that doesn't really take into consideration daily
   variability. A single big spike could easily take us over the edge.
2. I'm working with numbers that are a bit off from the county numbers (I think).
3. You don't need to wait until the county enacts policy to act. There is no
   doubt that daily new cases are increasing at the fastest rate since the start
   of this pandemic.

**Even more importantly:**

1. Under the **CURRENT** restrictions, gatherings of multiple households must be
  **outside**, with no more than **15 people**, from a maximum of
  **3 households**.
2. Be careful after Thanksgiving because many families will not follow these
   restrictions, which means in the immediate days/weeks following Thanksgiving
   there will likely be big increases in rates of spread and this will not be
   immediately obvious in reported numbers.

Stay safe out there!
